PhD Thesis

### 2016-12-04 pldi14/sec-change-theory.tex
Revise PLDI14 description of changes
Updating that text (with its clear structure) is currently easier, and
that text has more content. In particular, we already have text that
lets us talk about derivatives.

  \begin{subdefinition}
  \item $V$ is a set, called the \emph{base set}.
  \item Given $v \in V$, $\Change{v}$ is a set, called the \emph{change set}.
  \item Given $v \in V$ and $\D v \in \Change{v}$, $\Apply{\D v}{v} \in V$.
    \label{def:update}
  \item Given $u, v \in V$, $\Diff{u}{v} \in \Change{v}$.
    \label{def:diff}
  \item Given $u, v \in V$, $\Apply{\Diff*{u}{v}}{v}$ equals $u$.
    \qed
    \label{def:update-diff}
  \end{subdefinition}
\end{definition}

One might expect a further assumption that
$\Diff{\Apply*{\D v}{v}}{v} = \D v$. While it does hold
for the change structure of $\mathbb{N}$, it is not needed in general.
This means that multiple changes can represent the difference between
the same two base values. Throughout our theory, we only discuss equality of
base values, not of changes.

@@

@@

\paragraph{Nil changes and derivatives}
A particularly important change is the \emph{nil change} of a value:
\begin{definition}[Nil change]
  \label{def:nil-change}
  Given a change structure $\ChangeStruct{V}$ and a value $v \in V$, the change
  $\Diff{v}{v}$ is the nil change for $v$.
  \[
    \NilC{v} = \Diff{v}{v} \qed
  \]
\end{definition}
The nil change for a value does indeed not change it.
\begin{lemma}[Behavior of $\NILC$]
  \label{thm:update-nil}
  Given a change structure $\ChangeStruct{V}$ and a value $v \in V$,
  $\Apply{\NilC{v}}{v} = v$.
\end{lemma}

\begin{optionalproof}
Follows from \cref{def:update-diff,def:nil-change}.
\end{optionalproof}

\pg{Maybe should move this before nil changes?}
After defining change structures, we can restate the definition of derivatives from \cref{eq:correctness}.

\begin{definition}[Derivatives]
  \label{def:derivatives}
  Given change structures $\ChangeStruct{A}$ and $\ChangeStruct{B}$ and a function $f \in A \to
  B$ on the change sets of these change structures, we call a binary function $f'$ the \emph{derivative} of $f$ if
  for all values $a \in A$ and corresponding changes $\D a \in
  \Change[A]{a}$,
  \[\App{f}{\Apply*{\D a}{a}} = \Apply{\App{\App{f'}{a}}{\D a}}{\App{f}{a}}\text{.}\qed\]
\end{definition}

Applying a derivative to a value and its nil change gives a nil
change.%
\footnote{Post-print note: There's a small technical mistake in
  the following lemma. See \cref{sec:change-eq} for a corrected
  statement.}
%
\begin{lemma}[Behavior of derivatives on $\NILC$]
  \label{thm:deriv-nil}
  Given change structures $\ChangeStruct{A}$ and
  $\ChangeStruct{B}$, a function $f \in A \to B$, an element $a$
  of $A$, and the derivative $f'$ of $f$, we have
  $\App{\App{f'}{a}}{\NilC{a}} = \NilC{\App* f a}$.
\end{lemma}

>>>>

  \begin{subdefinition}
  \item $V$ is a set, called the \emph{base set}.
  \item Given $v \in V$, $\Change{v}$ is a set, called the \emph{change set}.
  \item Given $v \in V$ and $\D v \in \Change{v}$,
    $\Apply{\D v}{v} \in V$. We also call $v$ the \emph{base
      value} of the change, or its \emph{source}; we call
    $\Apply{\D v}{v}$ the \emph{updated value} of the change, or
    its \emph{destination}/\emph{target}.
    \label{def:update}
  \item Given $u, v \in V$, $\Diff{u}{v} \in \Change{v}$.
    \label{def:diff}
  \item Given $u, v \in V$, $\Apply{\Diff*{u}{v}}{v}$ equals $u$.
    \qed
    \label{def:update-diff}
  \end{subdefinition}
\end{definition}

@@

\paragraph{Change equivalence}
One might expect, in the definition of change structures
(\cref{def:change-struct}), the additional assumption that
$\Diff{\Apply*{\D v}{v}}{v} = \D v$. While this assumption holds
for the change structure of $\mathbb{N}$, there are sensible
examples of change structures where this assumption is not valid.
In fact, we can have multiple changes between the same source and
target. \pg{Add example on lists.} Therefore, we define an
equivalence among such changes that we call \emph{change
  equivalence}. When it is clear we are talking about changes, we
will also say that two changes are equivalent to mean that they
are change-equivalent. The definition of change equivalence is as
follows:

\begin{definition}[Change equivalence]
  Given a change structure $\ChangeStruct{V}$, a value $v \in V$,
  and two changes $\D v_1, \D v_2$ having $v$ as source, that is,
  such that $\D v_1, \D v_2\in \Change{v}$, we say that $\D v_1$
  is change-equivalent to $\D v_2$ and write
  $\D v_1 \Doe \D v_2$, if and only if these changes share,
  beyond the source $v$, also their target, that is, if and only
  if $\Update{v}{\D v_1} = \Update{v}{\D v_2}$.
\end{definition}

\begin{lemma}
  \label{def:diff-update-lemma}
  Given a change structure |chs(V) = (V, Dt, `oplus`, `ominus`)|,
  for any base value |v `elem` V| and for any change |dv| valid
  for |v| (that is, |dv `elem` Dt v| \pg{readd ^ to Dt}), we have |(v `oplus` dv)
  `ominus` v `doe` dv|.
\end{lemma}
\begin{proof}
Since both sides are changes for |v|, the thesis is equivalent to |v `oplus` ((v
`oplus` dv) `ominus` v) = v `oplus` dv|.

To prove our thesis, we remember that thanks to \cref{def:update-diff},
for any |v1, v2 : V| we have |v1 `oplus` (v2 `ominus` v1) = v2|. We can take |v1
= v|, |v2 = v `oplus` dv| and obtain |v `oplus` ((v `oplus` dv) `ominus` v) = v
`oplus` dv|, which is exactly our thesis.
\end{proof}

Change equivalence is indeed an equivalence relation, as stated
in the following lemma:
\begin{lemma}[Change equivalence is indeed an equivalence relation]
  For any change structure |chs(V)| and for any base value |v
  `elem` V|, change equivalence is an equivalence relation
  (reflexive, symmetric, transitive) among elements of |Dt v|.
\end{lemma}

As we will see, each valid operations in our theory will respect
change equivalence: equivalent changes will be mapped to
equivalent changes or to equal values.
\pg{Add example: however, different changes might have different values.}

@@

\subsubsection{Derivatives}
After defining change structures, we can define more formally
derivatives of functions, using a variant of
\cref{eq:correctness}.

\begin{definition}[Derivatives]
  \label{def:derivatives}
  Given change structures $\ChangeStruct{A}$ and $\ChangeStruct{B}$ and a function $f \in A \to
  B$ on the change sets of these change structures, we call a binary function $f'$ a \emph{derivative} of $f$ if
  for all values $a \in A$ and corresponding changes $\D a \in
  \Change[A]{a}$,
  \[\App{f}{\Apply*{\D a}{a}} = \Apply{\App{\App{f'}{a}}{\D a}}{\App{f}{a}}\text{.}\qed\]
\end{definition}

Here we are defining derivatives of mathematical functions, but
we will use them to define the meaning of derivatives of
programs. Intuitively, once we define a suitable set-theoretic
denotational semantics for programs, and a program transformation
that takes a program |f| to its derivative |f'|, we will ensure
that our semantics takes a program derivative |[[f']]| to a
derivative of the semantics of the base program |[[f]]|.

\pg{Point out here that derivatives respect change equivalence.}
\subsubsection{Nil changes and derivatives}
A change can have equal source and target, in which case we call
it a \emph{nil change}. We use $\DIFF$ to associate, to each
value, a distinguished nil change.
\pg{Rewrite}
\begin{definition}[Nil change]
  \label{def:nil-change}
  Given a change structure $\ChangeStruct{V}$ and a value $v \in V$, the change
  $\Diff{v}{v}$ is the nil change for $v$.
  \[
    \NilC{v} = \Diff{v}{v} \qed
  \]
\end{definition}
A nil change for a value does indeed not change it.
\begin{lemma}[Behavior of $\NILC$]
  \label{thm:update-nil}
  Given a change structure $\ChangeStruct{V}$ and a value $v \in V$,
  $\Apply{\NilC{v}}{v} = v$.
\end{lemma}

\begin{proof}
Follows from \cref{def:update-diff,def:nil-change}.
\end{proof}

Applying a derivative to a value and its nil change gives a nil
change.
%
\begin{lemma}[Behavior of derivatives on $\NILC$]
  \label{thm:deriv-nil}
  Given change structures $\ChangeStruct{A}$ and
  $\ChangeStruct{B}$, a function $f \in A \to B$, an element $a$
  of $A$, and the derivative $f'$ of $f$, we have
  $\App{\App{f'}{a}}{\NilC{a}} \Doe \NilC{\App* f a}$.
\end{lemma}
